---
title: "Flight Delay"
author: "Simge Cinar & Hugo Morvan"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Flight Delay}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r library}
library(ridgereg)
```

## QUESTION 1.2.1:

### 1) Read in the weather dataset and the flights dataset, remove eventual variables you do not believe to have a predictive value.
```{r}
data(flights, package = "nycflights13")
data(weather, package = "nycflights13")
```

```{r}
library(dplyr)
flights <- flights %>%
  mutate(new_time_origin = paste(origin, year, month, day, hour, sep = "_"))
flights <- flights[,c("new_time_origin", "arr_delay", "origin", "year", "month")]

weather <- weather %>%
  mutate(new_time_origin = paste(origin, year, month, day, hour, sep = "_"))
weather <- weather[,c("new_time_origin", "temp", "humid", "wind_speed", "pressure", "visib")]

merged_data <- flights %>%
  left_join(weather, by = "new_time_origin")

merged_data <- na.omit(merged_data) # remove the rows that contain NA
```

### 2) Add extra weather data from the weather dataset and create interaction effects you think could be of interest for the prediction.

```{r}
merged_data <- merged_data %>%
  mutate(temp_month_interaction = temp * month,
         temp_humid_interaction = temp * humid,
         temp_wind_interaction = temp * wind_speed,
         wind_humid_interaction = humid * wind_speed,
         wind_pressure_interaction = pressure * wind_speed,
         humid_visib_interaction = humid * visib)
```

```{r}
head(merged_data)
```


### 3) Divide the flight dataset into three sets: test, train and validation (with the proportions 5%, 80% and 15%.
```{r}
#library(caret)
set.seed(42)

# Create an initial split for training (80%) and the remaining data (20%)
initial_split <- caret::createDataPartition(merged_data$arr_delay, p = 0.8, list = FALSE)
remaining_data <- merged_data[-initial_split, ]

# Create a second split for validation (15%) from the remaining data
validation_split <- caret::createDataPartition(remaining_data$arr_delay, p = 0.75, list = FALSE)

train_data <- merged_data[initial_split, ]
validation_data <- remaining_data[validation_split, ]
test_data <- remaining_data[-c(validation_split), ]

row.names(train_data) <- NULL
row.names(validation_data) <- NULL
row.names(test_data) <- NULL

cat("Number of rows in the train data:", nrow(train_data), "\n")
cat("Number of rows in the validation data:", nrow(validation_data), "\n")
cat("Number of rows in the test data:", nrow(test_data), "\n")
```

### 4) Train ridge regressions models for different values of λ and evaluate the root mean squared error on the validation set. Try to find an optimal value for λ.

Model 1:
```{r}
formula <- arr_delay ~ temp + wind_speed
data <- train_data
lambda <- c(10^6, 10^5, 10^4, 10^3, 10^2, 10^1, 1, 10^-1, 10^-2, 10^-3, 10^-4, 10^-5, 10^-6)
rmse_val <- c()
for (l in lambda){
  ridgereg_model <- ridgereg:::ridgereg$new(formula, data, l)
  val_mat <- model.matrix(formula, validation_data)
  res <- ridgereg_model$predict(val_mat) - validation_data$arr_delay
  rmse <- sqrt(sum(res^2/nrow(validation_data)))
  rmse_val <- c(rmse_val, rmse)
}

cat("Best lambda is ", lambda[which.min(rmse_val)], "with", min(rmse_val), "RMSE")
```

Model 2:
```{r}
formula <- arr_delay ~ temp + wind_speed + humid 
data <- train_data
lambda <- c(10^6, 10^5, 10^4, 10^3, 10^2, 10^1, 1, 10^-1, 10^-2, 10^-3, 10^-4, 10^-5, 10^-6)
rmse_val <- c()
for (l in lambda){
  ridgereg_model <- ridgereg:::ridgereg$new(formula, data, l)
  val_mat <- model.matrix(formula, validation_data)
  res <- ridgereg_model$predict(val_mat) - validation_data$arr_delay
  rmse <- sqrt(sum(res^2/nrow(validation_data)))
  rmse_val <- c(rmse_val, rmse)
}

cat("Best lambda is ", lambda[which.min(rmse_val)], "with", min(rmse_val), "RMSE")
```

Model 3:
```{r}
formula <- arr_delay ~ temp + wind_speed + humid + visib
data <- train_data
lambda <- c(10^6, 10^5, 10^4, 10^3, 10^2, 10^1, 1, 10^-1, 10^-2, 10^-3, 10^-4, 10^-5, 10^-6)
rmse_val <- c()
for (l in lambda){
  ridgereg_model <- ridgereg:::ridgereg$new(formula, data, l)
  val_mat <- model.matrix(formula, validation_data)
  res <- ridgereg_model$predict(val_mat) - validation_data$arr_delay
  rmse <- sqrt(sum(res^2/nrow(validation_data)))
  rmse_val <- c(rmse_val, rmse)
}

cat("Best lambda is ", lambda[which.min(rmse_val)], "with", min(rmse_val), "RMSE")
```

Model 4:
```{r}
formula <- arr_delay ~ temp + wind_speed + humid + visib + month + origin
data <- train_data
lambda <- c(10^6, 10^5, 10^4, 10^3, 10^2, 10^1, 1, 10^-1, 10^-2, 10^-3, 10^-4, 10^-5, 10^-6)
rmse_val <- c()
for (l in lambda){
  ridgereg_model <- ridgereg:::ridgereg$new(formula, data, l)
  val_mat <- model.matrix(formula, validation_data)
  res <- ridgereg_model$predict(val_mat) - validation_data$arr_delay
  rmse <- sqrt(sum(res^2/nrow(validation_data)))
  rmse_val <- c(rmse_val, rmse)
}

cat("Best lambda is ", lambda[which.min(rmse_val)], "with", min(rmse_val), "RMSE")
```

Model 5:
```{r}
formula <- arr_delay ~ temp + wind_speed + humid + visib + month + origin + temp_wind_interaction
data <- train_data
lambda <- c(10^6, 10^5, 10^4, 10^3, 10^2, 10^1, 1, 10^-1, 10^-2, 10^-3, 10^-4, 10^-5, 10^-6)
rmse_val <- c()
for (l in lambda){
  ridgereg_model <- ridgereg:::ridgereg$new(formula, data, l)
  val_mat <- model.matrix(formula, validation_data)
  res <- ridgereg_model$predict(val_mat) - validation_data$arr_delay
  rmse <- sqrt(sum(res^2/nrow(validation_data)))
  rmse_val <- c(rmse_val, rmse)
}

cat("Best lambda is ", lambda[which.min(rmse_val)], "with", min(rmse_val), "RMSE")
```

Model 6:
```{r}
formula <- arr_delay ~ temp + wind_speed + humid + visib + month + origin + temp_humid_interaction
data <- train_data
lambda <- c(10^6, 10^5, 10^4, 10^3, 10^2, 10^1, 1, 10^-1, 10^-2, 10^-3, 10^-4, 10^-5, 10^-6)
rmse_val <- c()
for (l in lambda){
  ridgereg_model <- ridgereg:::ridgereg$new(formula, data, l)
  val_mat <- model.matrix(formula, validation_data)
  res <- ridgereg_model$predict(val_mat) - validation_data$arr_delay
  rmse <- sqrt(sum(res^2/nrow(validation_data)))
  rmse_val <- c(rmse_val, rmse)
}

cat("Best lambda is ", lambda[which.min(rmse_val)], "with", min(rmse_val), "RMSE")
```

Model 7:
```{r}
formula <- arr_delay ~ temp + wind_speed + humid + visib + month + origin + wind_humid_interaction
data <- train_data
lambda <- c(10^6, 10^5, 10^4, 10^3, 10^2, 10^1, 1, 10^-1, 10^-2, 10^-3, 10^-4, 10^-5, 10^-6)
rmse_val <- c()
for (l in lambda){
  ridgereg_model <- ridgereg:::ridgereg$new(formula, data, l)
  val_mat <- model.matrix(formula, validation_data)
  res <- ridgereg_model$predict(val_mat) - validation_data$arr_delay
  rmse <- sqrt(sum(res^2/nrow(validation_data)))
  rmse_val <- c(rmse_val, rmse)
}

cat("Best lambda is ", lambda[which.min(rmse_val)], "with", min(rmse_val), "RMSE")
```

Model 8:
```{r}
formula <- arr_delay ~ temp + wind_speed + humid + visib + month + origin + temp_wind_interaction + temp_humid_interaction + wind_humid_interaction
data <- train_data
lambda <- c(10^6, 10^5, 10^4, 10^3, 10^2, 10^1, 1, 10^-1, 10^-2, 10^-3, 10^-4, 10^-5, 10^-6)
rmse_val <- c()
for (l in lambda){
  ridgereg_model <- ridgereg:::ridgereg$new(formula, data, l)
  val_mat <- model.matrix(formula, validation_data)
  res <- ridgereg_model$predict(val_mat) - validation_data$arr_delay
  rmse <- sqrt(sum(res^2/nrow(validation_data)))
  rmse_val <- c(rmse_val, rmse)
}

cat("Best lambda is ", lambda[which.min(rmse_val)], "with", min(rmse_val), "RMSE")
```


### 5) When you found a good value for λ, use this to predict the test set and report the RMSE of your predicted model.
```{r}
formula <- arr_delay ~ temp + wind_speed + humid + temp_wind_interaction + temp_humid_interaction + wind_humid_interaction
data <- train_data

ridgereg_model <- ridgereg:::ridgereg$new(formula, data, 100)
test_mat <- model.matrix(formula, test_data)

res <- ridgereg_model$predict(test_mat) - test_data$arr_delay
rmse <- sqrt(sum(res^2/nrow(test_data)))
cat("RMSE on the test data:", rmse)
```

```{r}
library(ggplot2)
actual_values <- test_data[["arr_delay"]]
predicted_values <- as.vector(ridgereg_model$predict(test_mat))
data <- data.frame(Actual = actual_values, Predicted = predicted_values)

# Create a scatterplot
ggplot2::ggplot(data, aes(x = Actual, y = Predicted)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1, color = "blue", linetype = "dashed") +
  labs(x = "Actual Values", y = "Predicted Values") +
  ggtitle("Actual vs. Predicted Values")
```









